Molecular transport in human pial perivasculature

### 2024-11-27 paper/supplementary_main.tex
Updates from Overleaf
Former-commit-id: 4979054c206f9013549cfe20f7a45416e5c88b29

To complete~\eqref{eq:1d_pde}, we prescribe bifurcation conditions at
internal nodes of the perivascular networks $\Lambda_a, \Lambda_v$ and
boundary conditions at the end nodes. To this end, define the set of
internal nodes $\mathcal{B} \subset \mathcal{V}$ as the set of nodes
that are connected to two or more edges and the set of end nodes as
$\mathcal{T}$, write $\hat{c}^i = \hat{c}|_{\Lambda^i}$, and let
$\mathcal{E}(y)$ denote the set of (one, two or three) edges in
$\{\Lambda^i\}$ sharing the node $y$. To impose continuity of the
concentrations, we set that the concentrations when viewed from each
edge must match at nodes:
\begin{equation}
  \hat{c}^i (y) = \hat{c}^j(y)
  \qquad \foralls \Lambda^i, \Lambda^j \in \mathcal{E}(y)
  \qquad \foralls y \in \mathcal{B} .
  \label{eq:conc_continuity}
\end{equation}
Moreover, to ensure \discuss{mass conservation}, we set that the flux
going in and out at nodes should add to zero, or more precisely, that
for each $y \in \mathcal{B}$:
\begin{equation}
  \sum_{\Lambda^i \in \mathcal{E}(y)}
  \left (\hat{D}^i A^i \partial_s \hat{c}^i (y) - A^i \hat u^i(y) \hat{c}^i(y) \right ) n_i(y)
  = 0
  \label{eq:flux_conservation}
\end{equation}
where the function $n^i : \mathcal{V} \rightarrow \{-1, 0, 1\}$
defines an orientation of each vertex relative to the
edges. Specifically, for an edge $\Lambda^i$ from $y_{\rm in}$ to $y_{\rm out}$
\begin{equation}
  n^i(y_{\rm in}) = 1, \quad
  n^i(y_{\rm{out}}) = -1, \quad
  n^i(y) = 0 \quad \foralls y \in 
  \mathcal{V} \backslash \{ y_{\rm in}, y_{\rm out} \} .  
  \label{eq:def:sign}
\end{equation}

@@

\subsubsection{Modelling CSF flow via the incompressible Stokes equations}

The time-independent, incompressible Stokes equations model the flow
of an incompressible Newtonian fluid at low Reynolds numbers, and read
as follows: over the domain $\Omega_{\rm CSF} \subset \R^3$, find the
velocity vector field $\bm u : \Omega_{\rm CSF} \rightarrow \R^3$ and
the pressure field $p : \Omega_{\rm CSF} \rightarrow \R$ such that
\begin{subequations}
  \begin{alignat}{2}
    - \mu \Delta \bm u  + \nabla p &=  0 \quad && \text{ in } \Omega_{\rm CSF},
    \label{eq:momnetum_equation} \\ 
    \nabla \cdot \bm u &= 0 \quad && \text{ in } \Omega_{\rm CSF},
    \label{eq:divergence_equation} .
  \end{alignat}
  \label{eq:stokes}
\end{subequations}
In addition, we impose the following boundary conditions to model flow
induced by CSF production in the choroid plexus with $\Gamma_{\rm AM-U}$ as the main efflux site:
\begin{subequations}
  \begin{alignat}{2}
    \mu \nabla \bm u \cdot \bm{n} - p \bm n &= -R_0 ( \bm u \cdot \bm n ) \bm n
    && \text{ on } \Gamma_{\mathrm{AM-U}}, \\
    \bm u &= 0 && \text{ on } \Gamma_{\mathrm{AM-L}}, \\
    \bm u &= 0 && \text{ on } \Gamma_{\rm{pia}}, \\
    \bm u &= 0 && \text{ on } \Gamma_{\rm{SSAS}}, \\
    \bm u \cdot \bm n & = \frac{1}{ |\Gamma_{\rm LV}|}  u_{\rm in}, \quad \bm u \cdot \bm \tau = 0 \quad && \text{ on } \Gamma_{\rm{LV}} ,  
  \end{alignat}
  \label{eq:stokes_bcs}
\end{subequations}
where $\bm n$ denotes the unit outward normal to the boundary, $\mu$ is the (dynamic) CSF viscosity, $R_0 > 0$ is a resistance parameter for CSF efflux, and $u_{\rm in}$ is a given fluid influx, here across the lateral ventricle wall $\Gamma_{\rm LV}$. We consider alternative variations of these boundary conditions in connection with estimating dispersion effects induced by CSF pulsatility, see~\Cref{sec:supp:dispersion}.

@@

We first consider the discretization of~\eqref{eq:3d_pde} and
introduce the discrete space
\begin{equation}
  %V_h^k = \{c_h \in L^2(\Omega): \;  v_h \vert_E \in \mathbb{P}^k(E), \; E \in \mathcal{T}_h\}.
  V_h^k = \{c \in L^2(\discuss{\mathcal{T}}): \;  c \vert_E \in \mathbb{P}^k(E), \; E \in \mathcal{T}\} ,
\end{equation}
where $L^2(\mathcal{T})$ is the space of square-integrable functions
defined relative to $\mathcal{T}$ and $\mathbb{P}^k(E)$ denotes the
space of polynomials of degree $\leqslant k$ defined over the
tetrahedra $E$. To discretize the diffusion term in~\eqref{eq:3d_pde},
we use a symmetric weighted interior penalty DG formulation, referring
to~\cite{ern2009discontinuous} and \cite[Section
  4.5.2.3]{di2011mathematical} for details on this
method. \discuss{Recall that $\phi$ is constant in each domain
  $\Omega_{\rm PAR}$ and $\Omega_{\rm CSF}$, and thus in particular
  that $\nabla (\phi c) = \phi \nabla c$ on each $E \in \mathcal{T}$.}
Define for $c, v \in V_h^k$:
\begin{equation}
  a_h (c, v)
  =
  \sum_{E \in \mathcal{T}} \int_{E} D \discuss{\phi} \nabla c \cdot \nabla v
  + \sum_{F \in \mathcal{F}_i} \eta \frac{\gamma_{D,F}}{h_F} \int_{F} [c][v] 
  - \sum_{F \in \mathcal{F}_i} \int_{F} \left( \{D \phi \nabla c \}_w \cdot \bm n_F [v]
  + \{D \phi \nabla v \}_w \cdot \bm n_F [c] \right).
  \label{eq:dg_sip}
\end{equation}
In~\eqref{eq:dg_sip}, for each facet $F \in \mathcal{F}^i$ shared
between cells $E^1$ and $E^2$ with facet normal vector $\bm n_F$
pointing from $E^1$ to $E^2$, \discuss{$h_F$ denotes the facet size},
the jump $[\cdot]$ is given by $[v] = v|_{E^1} - v|_{E^2}$, and the
unweighted average $\{ \cdot \}$ and weighed average $\{ \cdot \}_w$
are defined as:
\begin{equation}
  \{v\} = \frac12 v |_{E^1} + \frac12 v |_{E^2}, \qquad
  \{v\}_{w} = \frac{\kappa_2}{\kappa_1 + \kappa_2} v |_{E^1} + \frac{\kappa_1}{\kappa_1 + \kappa_2} v |_{E^2}
  \quad \text{where} \quad \kappa_i = \phi D|_{E^i}.  
\end{equation}
The parameter $\gamma_{D,F}$ is the harmonic mean of the
\discuss{porosity-weighted} diffusion coefficient given by
\begin{equation}
  \gamma_{D,F} = \frac{\kappa_1\kappa_2}{\kappa_1 + \kappa_2},
\end{equation}
and $\eta$ is a user-specified penalty parameter. To discretize the convection term in~\eqref{eq:3d_pde}, we use
upwinding, see \cite[Section 2.3.1]{di2011mathematical} and the
references therein. For $c, v \in V_h^k$, define
\begin{equation}
  a_h^{\mathrm{up}}(c, v)
  = - \sum_{E \in \mathcal{T}} \int_E \bm{u} c \nabla v
  + \sum_{F \in \mathcal{F}_i } \int_F \{ \bm u  c\} \cdot \bm{n}_F [v] +
  \sum_{F \in \mathcal{F}_i} \int_F \frac{|\bm u \cdot \bm n_F|}{2} [c] [v] .
\end{equation}
Our discrete formulation for \eqref{eq:3d_pde} with the interface conditions~\eqref{eq:3d_interface} and boundary conditions~\eqref{eq:3d_bcs} and given initial conditions $c_h^0$ then reads: for $n = 1, 2, \dots$, with $t^n - t^{n-1} = \tau$, find $c_h^n \in V_h^k$ such that for all $v \in V_h^k$:
\begin{equation}
  \int_{\Omega} \frac{1}{\tau} (\phi c_h^n - \phi c_h^{n-1}) v
  + a_h(c_h^n, v) + a_h^{\mathrm{up}}(c_h^n,v)
  + \int_{\Gamma_{\rm pia} \discuss{\cup \Gamma_{\rm LV}} } \beta_{\rm pia}[c_h^n][v] 
  + \int_{\Gamma_{\rm AM-U}} \beta_{\rm exit} c_h^n v 
  + \int_{\Lambda} \xi  P \left (\overline{c_h^n} - \hat c_h^n \right) \overline{v} 
  = 
  %\int_{\Omega} f^n v +
  \int_{\Gamma_{\rm \discuss{S}SAS}} g_{\rm influx} v
  %, \quad \forall v \in V_h^k
  .
\end{equation}

@@

We consider an H(div)-based finite element approximation of the incompressible Stokes equations~\eqref{eq:stokes} defined over $\Omega_{\rm CSF}$ with the boundary conditions~\eqref{eq:stokes_bcs}. Following\cite{hong2016robust}, we approximate the velocity field $\bm u$ and the pressure field $p$ with the following finite element spaces:
\begin{align*}
  \bm V_h &= \{ \bm v  \in H(\mathrm{div}, \Omega_{\mathrm{CSF}}): \;
  \bm v \vert_{E} \in \mathrm{BDM}^2(E), \; E \in \mathcal{T}_{\rm CSF}; \;
  \bm v \cdot \bm n = 0
  \text{ on } \Gamma_{\rm{pia}} \cup \Gamma_{\rm AM-L} \cup \Gamma_{\rm{SSAS}}, \;\; \bm v \cdot \bm n = \frac{1}{|\Gamma_{\mathrm{LV}}|} u_{\mathrm{in}} \text{ on } \Gamma_{\mathrm{LV}} \}  \\ 
  Q_h  &= \{q \in L^2(\Omega_{\rm CSF}): \; q \vert_E \in \mathbb{P}^1(E),
  \; E \in \mathcal{T}_{\rm CSF} \}. 
\end{align*}
Here, $H(\mathrm{div}, \Omega)$ is the space of $L^2(\Omega)$ vector fields with $L^2(\Omega)$ divergence, $\mathrm{BDM}^2$ is the Brezzi-Douglas--Marini element~\cite{brezzi1987mixed} of degree 2, and $\bm n$ and $\bm \tau$ are the unit outward normal and unit tangent vectors to each facet, respectively. Given any vector $\bm v$, the normal and tangential components on each facet are denoted and given by 
\begin{equation}
\bm v_n = (\bm v \cdot \bm n) \bm n, \quad \bm v_t = \bm v - \bm v_n. 
\end{equation}
Since $\bm V_h \subset H(\mathrm{div}, \Omega_{\rm CSF})$, then $[\bm v_n] = 0 $ on $\mathcal{F}_{\rm CSF}$. Continuity in the tangential component is enforced weakly via interior penalization. Define the form \mer{MER: Check $\Gamma_{\rm AM}$ part.}
\begin{multline}
  \mathcal{A}_h(\bm u, \bm v)
  = \sum_{E \in \mathcal{T}_{\rm CSF}} \int_{E} \mu \nabla \bm u : \nabla \bm v
  - \sum_{F \in \discuss{\mathcal{F}_{\rm CSF} \cup \mathcal{F}_{e}}} \left (
  \int_{F} \mu \{\nabla \bm u\} \cdot \bm n_F \cdot [\bm v_t] 
  - %\sum_{F \in \mathcal{F}_{h,c} \cup \mathcal{F}_{\Gamma}}
  \int_{F} \mu \{\nabla \bm v\}\cdot \bm n_F \cdot [\bm u_t]
  + %\sum_{F \in \mathcal{F}_{h,c} \cup \mathcal{F}_{\Gamma}}
  \frac{\sigma \mu}{h} \int_{F} [\bm u_t ] \cdot [\bm v_t] \right ),   
\end{multline}
where \discuss{$h$ denotes the element size}. We set the penalty parameter for the tangential continuity to be $\sigma = 20$. Our finite element discretization of the incompressible Stokes equations is then to find $(\bm u_h, p_h) \in \bm V_h \times Q_h$ such that 
\begin{subequations}
\begin{align}
  \mathcal{A}_h(\bm u_h, \bm v_h)
  + \sum_{F \in \mathcal{F}_{\rm AM-U}} \int_{F} R_0 (\bm u_h \cdot \bm n) (\bm v_h \cdot \bm n)
  - \int_{\Omega_{\rm CSF}} \nabla \cdot \bm v_h \, p_h  &= 0, \qquad \foralls \bm v_h \in \bm V_h, \\ 
\int_{\Omega_{\rm CSF}} \nabla \cdot \bm u_h \, q_h  &= 0, \qquad \foralls q_h \in Q_h.
\end{align}
\end{subequations}

@@

In addition, CSF pulsates in the SAS and ventricular system with each
cardiac and respiratory cycle.

Taking the large spatial differences of pulsatile CSF flow into account, we compute the local diffusion enhancement factor $R$ in the CSF-filled space $\Omega_{\rm CSF}$ by combining a computational model of cardiac induced CSF flow with theoretical estimates for shear-augmented (Taylor) dispersion \cite{taylor1953dispersion, watson1983diffusion}.

Sharp et al \cite{keith2019dispersion} estimated the enhancement of diffusion in the spinal subarachnoid space due to dispersive effects to average 5.8 times that of purely molecular diffusion with a theoretical maximum of $10^6$.

Similar to the case of CSF flow driven by production introduced in section \ref{sec:csf_fluid_vel}, we employ a Stokes flow model to predict CSF velocities at peak arterial dilation. In particular, we enforce a uniformly distributed inflow of $10$\,ml/s and $2$\,ml/s across the outer skull ($\Gamma_{\rm skull}$) and lateral ventricular surfaces ($\Gamma_{\rm LV}$) respectively, mimicking the effect of reduced CSF volume due to blood inflow. We prescribe a no-slip boundary condition on the pial surface $\Gamma_{\rm Pia}$ and zero-traction on the spinal boundary $\Gamma_{\rm SSAS}$, making the spinal compartment the only outflow route of CSF. Thus, we compute the cardiac-driven flow field by augmenting the Stokes flow model (eq. \ref{eq:momnetum_equation} and \ref{eq:divergence_equation}) with the following boundary conditions:

\begin{equation}\label{eq:cardiac_flow_bcs}
    \bm u = 0 \text{ on } \Gamma_{\rm Pia}; \quad 
    \mu \nabla \bm u \cdot \bm n - p \bm n = 0 \text{ on } \Gamma_{\rm SSAS}; \quad 
    \bm u = \frac{0.31 \cdot \bm n}{|\Gamma_{\rm LV}|} \,\text{ml/s} \text{ on } \Gamma_{\rm LV}; \quad
    \bm u = \frac{8 \cdot \bm n}{|\Gamma_{\rm skull}|} \,\text{ml/s} \text{ on } \Gamma_{\rm skull}.
\end{equation}

The obtained pressure field only account for viscous forces, albeit cardiac induced CSF flow is known to be unsteady and substantially impacted by transient inertial forces. Assuming an oscillatory flow with the heartbeat frequency $\omega = 2 \pi$, a mean gap width of the CSF-filled spaces of $2h=3$\,mm and CSF density $\rho=10^3\,\text{kg/m}^3$, the Womersley number $\alpha$ becomes $\alpha^2 = \frac{h^2 \omega \rho}{\mu} = 20.2$, which is unsteady. Based on theoretical considerations on the ratio of oscillatory flow to steady flow impedances in a tube, we account for inertial forces in the pulsatile flow by upscaling the computed viscous pressure field by a factor of $1 + \alpha^2 / 8$ (see \cite{van1998cardiovascular}, chapter. 4.3).

Further assuming unsteady dispersion, we follow Sharp et al \cite{keith2019dispersion} and calculate the enhancement factor $R$ from the non-dimensionalized pressure gradient $dP=\frac{|\nabla p|}{\omega \mu / h}$ \rami{Can we refer to the exact equation here from Sharp et al ? i find that they write $P = (\partial \tilde p / \partial \tilde x) / (\rho \omega \mu_e/h)$ ... so i am a bit confused
}as $R\approx dP^2 / \alpha^3$. Combining the computed pressure gradient with this scaling law yields a pointwise approximation of the dispersion factor. Finally, we account for the non-local nature of the dispersion mechanism by applying a smoothing technique to obtain a spatially varying, but locally averaged diffusion enhancement factor $R$ (see the visualization in \cref{fig:visualize_R} and \cref{sec:dispersion_details} for additional details).

%\subsubsection{Cardiac Dispersion Factor Estimation}
%\label{sec:dispersion_details}

To compute a spatially varying estimate of the diffusion enhancement factor $R$, we consider the Stokes flow problem (\cref{eq:momnetum_equation} and \cref{eq:divergence_equation}) equipped with the boundary conditions \cref{eq:cardiac_flow_bcs} describing cardiac induced CSF flow at peak arterial pulsation. We use the numerical method described in \cref{sec:details_numerical_method} to solve the Stokes system and compute the non-dimensionalized local pressure difference $dP=\frac{|\nabla p| \rho}{\omega \mu / h}$ pointwise. Here, $\mu$ is the CSF viscosity, $\omega = 2 \pi$ is the oscillatory heart beat frequency, $2h=3$\,mm the assumed gap width and $\rho=10^3\,\text{kg/m}^3$ the density of the CSF.

>>>>

To complete~\eqref{eq:1d_pde}, we prescribe bifurcation conditions at
internal nodes of the perivascular networks $\Lambda_a, \Lambda_v$ and
boundary conditions at the end nodes. To this end, define the set of
internal nodes $\mathcal{B} \subset \mathcal{V}$ as the set of nodes
that are connected to two or more edges and the set of end nodes as
$\mathcal{T}$, write $\hat{c}^i = \hat{c}|_{\Lambda^i}$, and let
$\mathcal{E}(y)$ denote the set of (one, two or three) edges in
$\{\Lambda^i\}$ sharing the node $y$. To impose continuity of the
concentrations, we set that the concentrations when viewed from each
edge must match at nodes:
\begin{equation}
  \hat{c}^i (y) = \hat{c}^j(y)
  \qquad \foralls \Lambda^i, \Lambda^j \in \mathcal{E}(y)
  \qquad \foralls y \in \mathcal{B} .
  \label{eq:conc_continuity}
\end{equation}
Moreover, to ensure \discuss{mass conservation}, we set that the flux
going in and out at nodes should add to zero, or more precisely, that
for each $y \in \mathcal{B}$:
\begin{equation}
  \sum_{\Lambda^i \in \mathcal{E}(y)}
  \left (\hat{D}^i A^i \partial_s \hat{c}^i (y) - A^i \hat u^i(y) \hat{c}^i(y) \right ) ni(y)
  = 0
  \label{eq:flux_conservation}
\end{equation}
where the function $n^i : \mathcal{V} \rightarrow \{-1, 0, 1\}$
defines an orientation of each vertex relative to the
edges. Specifically, for an edge $\Lambda^i$ from $y_{\rm in}$ to $y_{\rm out}$
\begin{equation}
  n^i(y_{\rm in}) = 1, \quad
  n^i(y_{\rm{out}}) = -1, \quad
  n^i(y) = 0 \quad \foralls y \in 
  \mathcal{V} \backslash \{ y_{\rm in}, y_{\rm out} \} .  
  \label{eq:def:sign}
\end{equation}

@@

\subsubsection{Modelling CSF flow via the incompressible Stokes equations}
\label{sec:csf_flow}
The time-independent, incompressible Stokes equations model the flow
of an incompressible Newtonian fluid at low Reynolds numbers, and read
as follows: over the domain $\Omega_{\rm CSF} \subset \R^3$, find the
velocity vector field $\bm u : \Omega_{\rm CSF} \rightarrow \R^3$ and
the pressure field $p : \Omega_{\rm CSF} \rightarrow \R$ such that
\begin{subequations}
  \begin{alignat}{2}
    - \mu \Delta \bm u  + \nabla p &=  0 \quad && \text{ in } \Omega_{\rm CSF},
    \label{eq:momnetum_equation} \\ 
    \nabla \cdot \bm u &= 0 \quad && \text{ in } \Omega_{\rm CSF},
    \label{eq:divergence_equation} .
  \end{alignat}
  \label{eq:stokes}
\end{subequations}
In addition, we impose the following boundary conditions to model flow
induced by CSF production in the choroid plexus with $\Gamma_{\rm AM-U}$ as the main efflux site:
\begin{subequations}
  \begin{alignat}{2}
    \mu \nabla \bm u \cdot \bm{n} - p \bm n &= -R_0 ( \bm u \cdot \bm n ) \bm n
    && \text{ on } \Gamma_{\mathrm{AM-U}}, \\
    \bm u &= 0 && \text{ on } \Gamma_{\mathrm{AM-L}}\cup\Gamma_{\rm{pia}}\cup\Gamma_{\rm{SSAS}}, \\
    %\bm u &= 0 && \text{ on } \Gamma_{\rm{pia}}, \\
    %\bm u &= 0 && \text{ on } \Gamma_{\rm{SSAS}}, \\
    \bm u \cdot \bm n & = \frac{1}{ |\Gamma_{\rm LV}|}  u_{\rm in}, \quad \bm u \cdot \bm \tau = 0 \quad && \text{ on } \Gamma_{\rm{LV}} ,  
  \end{alignat}
  \label{eq:stokes_bcs}
\end{subequations}
where $\bm n$ denotes the unit outward normal to the boundary, $\mu$ is the (dynamic) CSF viscosity, $R_0 > 0$ is a resistance parameter for CSF efflux, and $u_{\rm in}$ is a given fluid influx, here across the lateral ventricle wall $\Gamma_{\rm LV}$. We consider alternative variations of these boundary conditions in connection with estimating dispersion effects induced by CSF pulsatility, see~\Cref{sec:supp:dispersion}.

@@

We first consider the discretization of~\eqref{eq:3d_pde} and
introduce the discrete space
\begin{equation}
  %V_h^k = \{c_h \in L^2(\Omega): \;  v_h \vert_E \in \mathbb{P}^k(E), \; E \in \mathcal{T}_h\}.
  V_h^k = \{c \in L^2(\discuss{\mathcal{T}}): \;  c \vert_E \in \mathbb{P}^k(E), \; E \in \mathcal{T}\} ,
\end{equation}
where $L^2(\mathcal{T})$ is the space of square-integrable functions
defined relative to $\mathcal{T}$ and $\mathbb{P}^k(E)$ denotes the
space of polynomials of degree $\leqslant k$ defined over the
tetrahedra $E$. To discretize the diffusion term in~\eqref{eq:3d_pde},
we use a symmetric weighted interior penalty DG formulation, referring
to~\cite{ern2009discontinuous} and \cite[Section
  4.5.2.3]{di2011mathematical} for details on this
method. \discuss{Recall that $\phi$ is constant in each domain
  $\Omega_{\rm PAR}$ and $\Omega_{\rm CSF}$, and thus in particular
  that $\nabla (\phi c) = \phi \nabla c$ on each $E \in \mathcal{T}$.}
Define for $c, v \in V_h^k$:
\begin{equation}
  a_h (c, v)
  =
  \sum_{E \in \mathcal{T}} \int_{E} D \discuss{\phi} \nabla c \cdot \nabla v
  + \sum_{F \in \mathcal{F}_i} \eta \frac{\gamma_{D,F}}{h_F} \int_{F} [c][v] 
  - \sum_{F \in \mathcal{F}_i} \int_{F} \left( \{D \phi \nabla c \}_w \cdot \bm n_F [v]
  + \{D \phi \nabla v \}_w \cdot \bm n_F [c] \right).
  \label{eq:dg_sip}
\end{equation}
In~\eqref{eq:dg_sip}, for each facet $F \in \mathcal{F}^i$ shared
between cells $E^1$ and $E^2$ with facet normal vector $\bm n_F$
pointing from $E^1$ to $E^2$, \discuss{$h_F$ denotes the facet diameter},
the jump $[\cdot]$ is given by $[v] = v|_{E^1} - v|_{E^2}$, and the
unweighted average $\{ \cdot \}$ and weighed average $\{ \cdot \}_w$
are defined as:
\begin{equation}
  \{v\} = \frac12 v |_{E^1} + \frac12 v |_{E^2}, \qquad
  \{v\}_{w} = \frac{\kappa_2}{\kappa_1 + \kappa_2} v |_{E^1} + \frac{\kappa_1}{\kappa_1 + \kappa_2} v |_{E^2}
  \quad \text{where} \quad \kappa_i = \phi D|_{E^i}.  
\end{equation}
The parameter $\gamma_{D,F}$ is the harmonic mean of the
\discuss{porosity-weighted} diffusion coefficient given by
\begin{equation}
  \gamma_{D,F} = \frac{\kappa_1\kappa_2}{\kappa_1 + \kappa_2},
\end{equation}
and $\eta$ is a user-specified penalty parameter. To discretize the convection term in~\eqref{eq:3d_pde}, we use
upwinding, see \cite[Section 2.3.1]{di2011mathematical} and the
references therein. For $c, v \in V_h^k$, define
\begin{equation}
  a_h^{\mathrm{up}}(c, v)
  = - \sum_{E \in \mathcal{T}} \int_E \bm{u} c \nabla v
  + \sum_{F \in \mathcal{F}_i } \int_F \{ \bm u  c\} \cdot \bm{n}_F [v] +
  \sum_{F \in \mathcal{F}_i} \int_F \frac{|\bm u \cdot \bm n_F|}{2} [c] [v] .
\end{equation}
Our discrete formulation for \eqref{eq:3d_pde} with the interface conditions~\eqref{eq:3d_interface} and boundary conditions~\eqref{eq:3d_bcs} and given initial conditions $c_h^0$ then reads: for $n = 1, 2, \dots$, with $t^n - t^{n-1} = \tau$, find $c_h^n \in V_h^k$ such that for all $v \in V_h^k$:
\begin{equation}
  \int_{\Omega} \frac{1}{\tau} (\phi c_h^n - \phi c_h^{n-1}) v
  + a_h(c_h^n, v) + a_h^{\mathrm{up}}(c_h^n,v)
  + \int_{\Gamma_{\rm pia} \discuss{\cup \Gamma_{\rm LV}} } \beta_{\rm pia}[c_h^n][v] 
  + \int_{\Gamma_{\rm AM-U}} \beta_{\rm exit} c_h^n v 
  + \int_{\Lambda} \xi  P \left (\overline{c_h^n} - \hat c_h^n \right) \overline{v} 
  = 
  %\int_{\Omega} f^n v +
  \int_{\Gamma_{\rm \discuss{S}SAS}} g_{\rm influx} v
  %, \quad \forall v \in V_h^k
  .
\end{equation}

@@

We consider an H(div)-based finite element approximation of the incompressible Stokes equations~\eqref{eq:stokes} defined over $\Omega_{\rm CSF}$ with the boundary conditions~\eqref{eq:stokes_bcs}. Following\cite{hong2016robust}, we approximate the velocity field $\bm u$ and the pressure field $p$ with the following finite element spaces:
\begin{align*}
  \bm V_h &= \{ \bm v  \in H(\mathrm{div}, \Omega_{\mathrm{CSF}}): \;
  \bm v \vert_{E} \in \mathrm{BDM}^2(E), \; E \in \mathcal{T}_{\rm CSF}; \;
  \bm v \cdot \bm n = 0
  \text{ on } \Gamma_{\rm{pia}} \cup \Gamma_{\rm AM-L} \cup \Gamma_{\rm{SSAS}}, \;\; \bm v \cdot \bm n = \frac{1}{|\Gamma_{\mathrm{LV}}|} u_{\mathrm{in}} \text{ on } \Gamma_{\mathrm{LV}} \}  \\ 
  Q_h  &= \{q \in L^2(\Omega_{\rm CSF}): \; q \vert_E \in \mathbb{P}^1(E),
  \; E \in \mathcal{T}_{\rm CSF} \}. 
\end{align*}
Here, $H(\mathrm{div}, \Omega)$ is the space of $L^2(\Omega)$ vector fields with $L^2(\Omega)$ divergence, $\mathrm{BDM}^2$ is the Brezzi-Douglas--Marini element~\cite{brezzi1987mixed} of degree 2, and $\bm n$ and $\bm \tau$ are the unit outward normal and unit tangent vectors to each facet, respectively. Given any vector $\bm v$, the normal and tangential components on each facet are denoted and given by 
\begin{equation}
\bm v_n = (\bm v \cdot \bm n) \bm n, \quad \bm v_t = \bm v - \bm v_n. 
\end{equation}
Since $\bm V_h \subset H(\mathrm{div}, \Omega_{\rm CSF})$, then $[\bm v_n] = 0 $ on $\mathcal{F}_{\rm CSF}$. Continuity in the tangential component is enforced weakly via interior penalization. Define the form \mer{MER: Check $\Gamma_{\rm AM}$ part. 26.11.24: Rami will check $\mathcal{F}_e$ versus definitions of the boundaries and where it appears.}
\begin{multline}
  \mathcal{A}_h(\bm u, \bm v)
  = \sum_{E \in \mathcal{T}_{\rm CSF}} \int_{E} \mu \nabla \bm u : \nabla \bm v
  - \sum_{F \in \discuss{\mathcal{F}_{\rm CSF} \cup \mathcal{F}_{e}}} \left (
  \int_{F} \mu \{\nabla \bm u\} \cdot \bm n_F \cdot [\bm v_t] 
  - %\sum_{F \in \mathcal{F}_{h,c} \cup \mathcal{F}_{\Gamma}}
  \int_{F} \mu \{\nabla \bm v\}\cdot \bm n_F \cdot [\bm u_t]
  + %\sum_{F \in \mathcal{F}_{h,c} \cup \mathcal{F}_{\Gamma}}
  \frac{\sigma \mu}{h} \int_{F} [\bm u_t ] \cdot [\bm v_t] \right ),   
\end{multline}
where \discuss{$h$ denotes the element size}. We set the penalty parameter for the tangential continuity to be $\sigma = 20$. Our finite element discretization of the incompressible Stokes equations is then to find $(\bm u_h, p_h) \in \bm V_h \times Q_h$ such that 
\begin{subequations}
\begin{align}
  \mathcal{A}_h(\bm u_h, \bm v_h)
  + \sum_{F \in \mathcal{F}_{\rm AM-U}} \int_{F} R_0 (\bm u_h \cdot \bm n) (\bm v_h \cdot \bm n)
  - \int_{\Omega_{\rm CSF}} \nabla \cdot \bm v_h \, p_h  &= 0, \qquad \foralls \bm v_h \in \bm V_h, \\ 
\int_{\Omega_{\rm CSF}} \nabla \cdot \bm u_h \, q_h  &= 0, \qquad \foralls q_h \in Q_h.
\end{align}
\end{subequations}

@@

CSF pulsates in the SAS and ventricular system with each cardiac and respiratory cycle. Taking the large spatial differences of pulsatile CSF flow into account, we estimate a local diffusion enhancement factor $R$ in the CSF-filled space $\Omega_{\rm CSF}$ by combining a computational model of cardiac- and respiration-induced pulsatile CSF flow with theoretical estimates for shear-augmented (Taylor) dispersion \cite{taylor1953dispersion, watson1983diffusion}.

\rami{I rewrote the above paragraph and some aspects of this section}

\rami{  CSF pulsates in the SAS and ventricular system with each cardiac and respiratory cycle with large spatial differences.  This leads to dispersion effects. For example, 
Sharp et al \cite{keith2019dispersion} estimated the enhancement of diffusion in the spinal subarachnoid space due to dispersive effects to average 5.8 times that of purely molecular diffusion with a theoretical maximum of $10^6$. 
This phenomena is accounted for in our model by scaling the free diffusion coeffiecient $D^{\mathrm{Gad}}$ by the formula $D^{\mathrm{eff}} = (1+R) D^{\mathrm{Gad}}$ where $R$ represents a local diffusion enhancement in the CSF-filled space $\Omega_{\rm CSF}$.}

\rami{ Note that since there is no data for the local values of $R$, we estimate $R$ by combining computational results (solving Stokes flow equations, see below) and explicit estimates for dispersion enhancement that are derived under simplifying assumptions on the problem geometry, the flow field and concentration profiles \cite{watson1983diffusion, keith2019dispersion}. Therefore such an estimate is heuristic; however,  our computational results qualitatively agree with observed data.  A key feature in our estimate of $R$ is that it is spatially varying, including local effects depending on a  computed flow field. }

Similar to the case of CSF flow driven by production introduced in \Cref{sec:csf_flow}, we employ a Stokes flow model to predict CSF velocities corresponding to the peak cardiac and respiratory blood inflow rates. In particular, we enforce a uniformly distributed inflow of $u^{\rm in}_{\rm AM}$ and $u^{\rm in}_{\rm LV}$ across the arachnoid mater ($\Gamma_{\rm AM}$) and lateral ventricular surfaces ($\Gamma_{\rm LV}$) respectively, mimicking the effect of reduced CSF volume due to blood inflow. We set $u^{\rm in}_{\rm AM} = 6\,$ml/s and $u^{\rm in}_{\rm LV}=0.31\,$ml/s in the cardiac case, and $u^{\rm{in}}_{\rm{AM}} = 1\,$ml/s and $u^{\rm in}_{\rm{LV}}=0.121\,$ml/s in the respiratory case, respectively.
We prescribe a no-slip boundary condition on the pial surface $\Gamma_{\rm Pia}$ and zero-traction on the spinal boundary $\Gamma_{\rm SSAS}$, making the spinal compartment the only outflow route of CSF. Thus, we compute the pulsatile flow fields by augmenting the Stokes flow model (eq. \ref{eq:momnetum_equation} and \ref{eq:divergence_equation}) with the following boundary conditions:
\begin{equation}\label{eq:cardiac_flow_bcs}
    \bm u = 0 \text{ on } \Gamma_{\rm Pia}; \quad 
    \mu \nabla \bm u \cdot \bm n - p \bm n = 0 \text{ on } \Gamma_{\rm SSAS}; \quad 
    \bm u = \frac{u^{\rm in}_{\rm LV} \cdot \bm n}{|\Gamma_{\rm LV}|} \text{ on } \Gamma_{\rm LV}; \quad
    \bm u = \frac{u^{\rm in}_{\rm AM} \cdot \bm n}{|\Gamma_{\rm skull}|} \text{ on } \Gamma_{\rm skull}.
\end{equation}
The obtained pressure fields $p_{\rm cardiac}$ and $p_{\rm resp}$ only account for viscous forces, albeit cardiac and respiratory induced CSF flow is known to be unsteady and substantially impacted by transient inertial forces. Assuming an oscillatory flow with either the heartbeat frequency $\omega_{\rm cardiac} = 2 \pi$ or respiratory frequency $\omega_{\rm resp} = 0.5 \pi$, a mean gap width of the CSF-filled spaces of $2h=3$\,mm and CSF density $\rho=10^3\,\text{kg/m}^3$, the Womersley numbers $\alpha$ become $\alpha_{\rm cardiac}^2 = \frac{h^2 \omega_{\rm cardiac} \rho}{\mu} = 20.2$ and $\alpha_{\rm resp}^2=5.05$, which is unsteady. Based on theoretical considerations on the ratio of oscillatory flow to steady flow impedances in a tube, we account for inertial forces in the pulsatile flow by upscaling the computed viscous pressure field by a factor of $1 + \alpha^2 / 8$ (see \cite{van1998cardiovascular}, chapter. 4.3). Thus, the upscaled pressure fields become $p^{\rm up}_{\rm cardiac}= (1 + \alpha_{\rm cardiac}^2 / 8)p$ and $p^{\rm up}_{\rm resp}= (1 + \alpha_{\rm resp}^2 / 8)p$.
